---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
  , comment = "#>"
  , fig.path = "man/figures/README-"
  , out.width = "100%"
  , echo = TRUE
  , warning = FALSE
  , message = FALSE
)
library(magrittr)
```

# cloud2trees

<!-- badges: start -->
<!-- badges: end -->

The goal of cloud2trees is to provide accessible routines for processing point cloud data collected by airborne lidar or developed using UAS imagery and photogrammetry (e.g. structure from motion). The cloud2trees package implements some of the methods outlined in the literature below. Some of the methodologies implemented in the cloud2trees package have been developed specifically to quantify *conifer* forest structure and may not be appropriate for other uses.

* Swayze, Neal C., and Wade T. Tinkham. ["Application of unmanned aerial system structure from motion point cloud detected tree heights and stem diameters to model missing stem diameters."](https://scholar.google.com/scholar?oi=bibs&hl=en&cluster=10655866445299954513) MethodsX 9 (2022): 101729.
* Tinkham, Wade T., Neal C. Swayze, Chad M. Hoffman, Lauren E. Lad, and Mike A. Battaglia. ["Modeling the missing DBHs: Influence of model form on UAV DBH characterization."](https://scholar.google.com/scholar?oi=bibs&hl=en&cluster=14807895789640069059) Forests 13, no. 12 (2022): 2077.
* Creasy, Matthew B., Wade T. Tinkham, Chad M. Hoffman, and Jody C. Vogeler. ["Potential for individual tree monitoring in ponderosa pine dominated forests using unmanned aerial system structure from motion point clouds."](https://scholar.google.com/scholar?oi=bibs&hl=en&cluster=10356932785437169630) Canadian Journal of Forest Research 51, no. 8 (2021): 1093-1105.
* Kane, Van R., Bryce N. Bartl-Geller, Gina R. Cova, Caden P. Chamberlain, Liz van Wagtendonk, and Malcolm P. North. ["Where are the large trees? A census of Sierra Nevada large trees to determine their frequency and spatial distribution across three large landscapes."](https://scholar.google.com/scholar?cluster=18100846251549158494&hl=en&as_sdt=0,6) Forest Ecology and Management 546 (2023): 121351.
* Almeida, Danilo Roberti Alves de, Scott C. Stark, Gang Shao, Juliana Schietti, Bruce Walker Nelson, Carlos Alberto Silva, Eric Bastos Gorgens, Ruben Valbuena, Daniel de Almeida Papa, and Pedro Henrique Santin Brancalion. ["Optimizing the remote detection of tropical rainforest structure with airborne lidar: Leaf area profile sensitivity to pulse density and spatial sampling."](https://scholar.google.com/scholar?cluster=5085674356305592014&hl=en&as_sdt=0,6) Remote Sensing 11, no. 1 (2019): 92.
* Viedma, O., C. A. Silva, J. M. Moreno, and A. T. Hudak. ["LadderFuelsR: A new automated tool for vertical fuel continuity analysis and crown base height detection using light detection and ranging."](https://scholar.google.com/scholar?cluster=14042621889986838235&oi=gsb&hl=en&as_sdt=0,6) Methods in Ecology and Evolution (2024).

## Installation

You must have Rtools to build packages in R. Check that you have it and install if not.

``` r
## pkgbuild helps us check for Rtools
install.packages("pkgbuild")

# check for Rtools which is required to build packages
pkgbuild::check_build_tools(debug = TRUE)
```

The `cloud2trees` package relies on the following packages which *must be manually installed first*:

* `lasR` package ([https://r-lidar.github.io/lasR/](https://r-lidar.github.io/lasR/)) helps us process point cloud data with speed
* `TreeLS` package ([https://github.com/tiagodc/TreeLS](https://github.com/tiagodc/TreeLS)) helps us extract DBH from the point cloud

``` r
## remotes helps us get packages hosted on github
install.packages("remotes")

## install lasR from the r-univers
install.packages("lasR", repos = "https://r-lidar.r-universe.dev")

## install TreeLS from github
remotes::install_github(repo = "tiagodc/TreeLS", upgrade = F)
```

Additionally, if you wish to estimate crown base height (CBH) as part of the point cloud processing the `LadderFuelsR` package ([https://github.com/olgaviedma/LadderFuelsR](https://github.com/olgaviedma/LadderFuelsR)) and `leafR` package ([https://github.com/DRAAlmeida/leafR](https://github.com/DRAAlmeida/leafR)) *must be manually installed first*.

``` r
## install LadderFuelsR
remotes::install_github(repo = "olgaviedma/LadderFuelsR", upgrade = F)
## install leafR
remotes::install_github(repo = "DRAAlmeida/leafR", upgrade = F)
```

Finally, you can install the development version of `cloud2trees` from [GitHub](https://github.com/georgewoolsey/cloud2trees) with:

```r
## get cloud2trees
remotes::install_github(repo = "georgewoolsey/cloud2trees", upgrade = F)
```

## Preliminaries

### tl;dr

The `cloud2trees` package relies on external data to estimate tree DBH and tree forest type. If you plan to use this functionality, the data must be downloaded first using `get_data()` which only needs to be run the first time you use the `cloud2trees` package.

```{r, results = FALSE}
library(cloud2trees)
# download the external data
cloud2trees::get_data()
```

Some machines, US Federal Government computers especially, might prevent you from writing directly to the package contents. If you get an error about "access" or "writing to" from the `get_data()` function, try to set the `savedir` parameter to a directory that you know you can write to.

```{r, eval=FALSE}
# download the external data with a custom directory
cloud2trees::get_data(savedir = "C:/ican/access/this")
```

We'll be using the `tidyverse`, `sf`, and `terra` in the examples below.

```{r}
# install.packages("tidyverse")
library(tidyverse)
# install.packages("sf")
library(sf)
# install.packages("terra")
library(terra)
```

### Get TreeMap Data

To estimate tree DBH from extracted tree height data requires training data to model diameter using height. Site-specific allometric equations using data from the USDA Forest Service’s Forest Inventory Analysis (FIA) program are used to estimate individual tree DBH based on the point cloud-detected tree height.

FIA plots are identified using [TreeMap 2016](https://scholar.google.com/scholar?cluster=10696389844245769168&hl=en&as_sdt=0,6), a model of FIA plot locations imputed throughout forested areas of the conterminous United States at 30 m spatial resolution.

Unless you have already executed the `get_data()` function, you must first download the FIA data (~3 GB) using `get_treemap()` which only needs to be run the first time you use the `cloud2trees` package.

```{r, results = FALSE}
# download the TreeMap data
cloud2trees::get_treemap()
```

### Get Forest Type Group Data

To estimate FIA forest type group for trees requires external data to spatially determine the type of a tree.

Forest type groups are identified using the [Forest Type Groups of the Continental United States](https://www.arcgis.com/home/item.html?id=10760c83b9e44923bd3c18efdaa7319d) data (Wilson 2023). This forest type group layer was developed using data from over 213,000 national forest inventory plots measured during the period 2014-2018 from the FIA program and has been aggregated to 90-meter resolution for this specific project to make the data more accessible over the entire continental US.

Unless you have already executed the `get_data()` function, you must first download the FIA data (~3 GB) using `get_foresttype()` which only needs to be run the first time you use the `cloud2trees` package.

```{r, results = FALSE}
# download the forest type data
cloud2trees::get_foresttype()
```

## Extract Trees from Point Cloud: Default

The `cloud2trees()` function is an all-in-one function to process raw .las|.laz files to generate a CHM raster (.tif), a DTM raster (.tif), and a tree list with tree location, height, and DBH.

For our example we'll use the `MixedConifer.laz` that ships with the `lidR` package ([https://r-lidar.github.io/lidRbook/](https://r-lidar.github.io/lidRbook/)).

A most basic example using all `cloud2trees()` function defaults with a single .laz file and writing the output to a temporary directory is:

```{r, results = FALSE}
# a test las file but this could also be a directory path with >1 .las|.laz files
i <- system.file("extdata", "MixedConifer.laz", package="lidR")
# run it
cloud2trees_ans <- cloud2trees::cloud2trees(output_dir = tempdir(), input_las_dir = i)
```

Let's check out what is included in the return from the `cloud2trees()` function.

```{r}
# what is it?
cloud2trees_ans %>% names()
```

There is a digital terrain model (DTM) raster which we can plot using `terra::plot()`

```{r}
# there's a DTM
cloud2trees_ans$dtm_rast %>% terra::plot()
```

There is a canopy height model (CHM) raster which we can plot using `terra::plot()`

```{r}
# there's a CHM
cloud2trees_ans$chm_rast %>% terra::plot()
```

A spatial data frame with tree crown polygons is returned.

```{r}
# there are tree crowns
cloud2trees_ans$crowns_sf %>% dplyr::glimpse()
```

Notice that all of the `dbh`, `cbh`, `forest_type`, HMD (`max_crown_diam_height_m`), and competition (`comp_`) columns do not have data. To estimate these values, we need to explicitly tell the `cloud2trees()` to perform the processing required by setting the parameters:

* `estimate_tree_dbh=TRUE` for DBH (see also [`trees_dbh()`](#trees_dbh))
* `estimate_tree_cbh=TRUE` for CBH (see also [`trees_cbh()`](#trees_cbh))
* `estimate_tree_type=TRUE` for forest type (see also [`trees_type()`](#trees_type))
* `estimate_tree_hmd=TRUE` for tree HMD (see also [`trees_hmd()`](#trees_hmd))
* `estimate_tree_competition=TRUE` for competition (see also `trees_competition()`)

Let's plot these tree crown polygons using `ggplot2::ggplot()` with some custom plot settings.

```{r}
cloud2trees_ans$crowns_sf %>% 
  ggplot2::ggplot(mapping = ggplot2::aes(fill = tree_height_m)) + 
  ggplot2::geom_sf() + 
  ggplot2::scale_fill_distiller(palette = "Oranges", name = "tree ht. (m)", direction = 1) +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "top", legend.direction = "horizontal")
```

A spatial data frame with tree top points is returned.

```{r}
# there are tree top points
cloud2trees_ans$treetops_sf %>% dplyr::glimpse()
```

Notice that `cloud2trees_ans$crowns_sf` and `cloud2trees_ans$treetops_sf` have the *exact same structure* but one is spatial polygons and the other is spatial points.

Let's plot these tree top points using `ggplot2::ggplot()` with some custom plot settings.

```{r}
cloud2trees_ans$treetops_sf %>% 
  ggplot2::ggplot(mapping = ggplot2::aes(color = tree_height_m)) + 
  ggplot2::geom_sf() + 
  ggplot2::scale_color_distiller(palette = "Oranges", name = "tree ht. (m)", direction = 1) +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "top", legend.direction = "horizontal")
```

It is also the case that the points in `cloud2trees_ans$treetops_sf` will match to exactly one crown polygon in `cloud2trees_ans$crowns_sf`.

```{r}
ggplot2::ggplot() + 
  ggplot2::geom_sf(data = cloud2trees_ans$crowns_sf, mapping = ggplot2::aes(fill = tree_height_m)) + 
  ggplot2::geom_sf(data = cloud2trees_ans$treetops_sf, shape = 20) + 
  ggplot2::scale_fill_distiller(palette = "Oranges", name = "tree ht. (m)", direction = 1) +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "top", legend.direction = "horizontal")
```

## Extract Trees from Point Cloud: Custom

We'll continue to use the `MixedConifer.laz` that ships with the `lidR` package for our example.

Customizing the `cloud2trees()` function parameters we'll:

* Change the resolution of the DTM using `dtm_res_m`
* Change the minimum height to classify trees using `min_height`
* Change moving window used to detect the local maxima tree tops using `ws`
* Estimate tree DBH using allometry from FIA plot data with `estimate_tree_dbh`
* Extract tree FIA Forest Type Group with `estimate_tree_type`
* Quantify tree competition metrics with `estimate_tree_competition`
* Extract tree CBH from the point cloud with `estimate_tree_cbh` for a sample of 333 trees using `cbh_tree_sample_n`
* Model the remaining tree CBH values with `cbh_estimate_missing_cbh` based on our sample of 333 trees

```{r, results = FALSE}
# run it
cloud2trees_ans_c <- cloud2trees::cloud2trees(
  output_dir = tempdir()
  , input_las_dir = i
  , dtm_res_m = 0.5
  , min_height = 3
  , ws = function(x){x*0.1}
  , estimate_tree_dbh = TRUE
  , estimate_tree_type = TRUE
  , estimate_tree_competition = TRUE
  , estimate_tree_cbh = TRUE
  , cbh_tree_sample_n = 333
  , cbh_estimate_missing_cbh = TRUE
)
```

Check how the digital terrain model (DTM) raster has changed

```{r}
paste(
  "Default DTM resolution:"
  , cloud2trees_ans$dtm_rast %>% terra::res() %>% paste(collapse = ",")
  , "|| Custom DTM resolution:"
  , cloud2trees_ans_c$dtm_rast %>% terra::res() %>% paste(collapse = ",")
)
```

Check that our spatial data frame with tree crown polygons has data in the `dbh`, `cbh`, `forest_type`, and `comp` (i.e. competition) columns.

```{r}
cloud2trees_ans_c$crowns_sf %>% dplyr::glimpse()
```

Remember, we also changed the `ws` parameter used to detect the local maxima for identifying tree tops so we got a lot more trees compared to the default settings.

```{r}
paste(
  "Default trees extracted:"
  , cloud2trees_ans$crowns_sf %>% nrow()
  , "|| Custom trees extracted:"
  , cloud2trees_ans_c$crowns_sf %>% nrow()
)
```

Let's look at the relationship between tree height and tree DBH estimated from the FIA plot data.

```{r}
cloud2trees_ans_c$crowns_sf %>%
  ggplot2::ggplot(mapping = ggplot2::aes(x = tree_height_m, y = dbh_cm)) + 
  ggplot2::geom_point(color = "navy", alpha = 0.6) +
  ggplot2::labs(x = "tree ht. (m)", y = "tree DBH (cm)") +
  ggplot2::scale_x_continuous(limits = c(0,NA)) +
  ggplot2::scale_y_continuous(limits = c(0,NA)) +
  ggplot2::theme_light()
```

Let's look at the relationship between tree height and tree CBH as extracted from the point cloud. Note, that we do not expect a perfect linear relationship between tree height and CBH throughout the entire height range because CBH is also determined spatially (e.g. as a fire moves through a stand).

```{r}
cloud2trees_ans_c$crowns_sf %>%
  dplyr::arrange(is_training_cbh) %>%
  ggplot2::ggplot(mapping = ggplot2::aes(x = tree_height_m, y = tree_cbh_m, color=is_training_cbh)) + 
  ggplot2::geom_point() +
  ggplot2::labs(x = "tree ht. (m)", y = "tree CBH (m)") +
  ggplot2::scale_y_continuous(breaks = scales::extended_breaks(n=12)) +
  ggplot2::scale_x_continuous(breaks = scales::extended_breaks(n=14)) +
  ggplot2::scale_color_viridis_d(alpha = 0.8, name = "is CBH\nfrom cloud?") +
  ggplot2::theme_light()
```

We can also plot height, diameter, and CBH of trees spatially and we'll use the `patchwork` package to combine our plots.

```{r}
library(patchwork)
# height plot
plt_ht <-
  cloud2trees_ans_c$crowns_sf %>% 
  ggplot2::ggplot(mapping = ggplot2::aes(fill = tree_height_m)) + 
  ggplot2::geom_sf() + 
  ggplot2::scale_fill_distiller(palette = "Oranges", name = "tree ht. (m)", direction = 1) +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "top", legend.direction = "horizontal")
# diameter plot
plt_dbh <-
  cloud2trees_ans_c$crowns_sf %>% 
  ggplot2::ggplot(mapping = ggplot2::aes(fill = dbh_cm)) + 
  ggplot2::geom_sf() + 
  ggplot2::scale_fill_distiller(palette = "Purples", name = "tree DBH (cm)", direction = 1) +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "top", legend.direction = "horizontal")
# CBH plot
plt_cbh <-
  cloud2trees_ans_c$crowns_sf %>% 
  ggplot2::ggplot(mapping = ggplot2::aes(fill = tree_cbh_m)) + 
  ggplot2::geom_sf() + 
  ggplot2::scale_fill_distiller(palette = "Greens", name = "tree CBH (m)", direction = 1) +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "top", legend.direction = "horizontal")
# combine with patchwork
plt_ht + plt_dbh + plt_cbh + patchwork::plot_layout(ncol = 2) &
  ggplot2::theme(
    legend.title = ggplot2::element_text(size = 8)
    , legend.text = ggplot2::element_text(size = 7)
  )
```

Let's plot the distance to the nearest tree that we obtained by turning on the `estimate_tree_competition` parameter in the `cloud2trees()` function call to quantify tree competition metrics. We'll use the spatial tree points data in `cloud2trees_ans_c$treetops_sf`. 

```{r}
cloud2trees_ans_c$treetops_sf %>%
  ggplot2::ggplot(mapping = ggplot2::aes(color = comp_dist_to_nearest_m)) + 
  ggplot2::geom_sf() +
  ggplot2::scale_color_distiller(palette = "Greys", name = "distance to\nnearest tree", direction = 1) +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "top", legend.direction = "horizontal")
```

Let's look at the FIA Forest Type Group data we extracted for the tree list.

```{r}
cloud2trees_ans_c$treetops_sf %>%
  sf::st_drop_geometry() %>% 
  dplyr::count(forest_type_group_code, forest_type_group)
```

## Extract Raster Data from Point Cloud

We can use the `cloud2raster()` function if we only want to create a DTM and CHM from our point cloud data. This function also creates a classified and height normalized point cloud in the process. If you wish to keep these point clouds ensure to turn on the `keep_intrmdt` parameter and see the `point_cloud_processing_temp` directory nested in the `output_dir`.

```{r, results = FALSE}
cloud2raster_ans <- cloud2trees::cloud2raster(output_dir = tempdir(), input_las_dir = i)
```

There is a digital terrain model (DTM) raster which we can plot using `terra::plot()`

```{r}
# there's a DTM
cloud2raster_ans$dtm_rast %>% terra::plot()
```

There is a canopy height model (CHM) raster which we can plot using `terra::plot()`

```{r}
# there's a CHM
cloud2raster_ans$chm_rast %>% terra::plot()
```

## Extract Trees from Raster Data

We can use the `raster2trees()` function if we already have a CHM raster and want to extract a tree list.

We'll use the CHM example that ships with the `cloud2trees` package.

```{r, results = FALSE}
# read example CHM raster
f <- paste0(system.file(package = "cloud2trees"),"/extdata/chm.tif")
r <- terra::rast(f)
# extract trees from raster
raster2trees_ans <- cloud2trees::raster2trees(chm_rast = r, outfolder = tempdir())
```

A spatial data frame with tree crown polygons is returned.

```{r}
# there are tree crowns
raster2trees_ans %>% dplyr::glimpse()
```

Let's plot these tree crown polygons using `ggplot2::ggplot()` with some custom plot settings.

```{r}
raster2trees_ans %>% 
  ggplot2::ggplot(mapping = ggplot2::aes(fill = tree_height_m)) + 
  ggplot2::geom_sf() + 
  ggplot2::scale_fill_distiller(palette = "Oranges", name = "tree ht. (m)", direction = 1) +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "top", legend.direction = "horizontal")
```

## Estimate Tree DBH for a Tree List{#trees_dbh}

If we already have a list of trees with tree coordinate and tree height data, we can estimate tree DBH using a site-specific allometric equation based on FIA data with the `trees_dbh()` function.

We just need to pass a data frame with the columns `treeID`, `tree_x`, `tree_y`, and `tree_height_m` to the `trees_dbh()` function.

```{r}
# a fake tree list
tl <- dplyr::tibble(
    treeID = c(1:21)
    , tree_x = rnorm(n=21, mean = 458064, sd = 11)
    , tree_y = rnorm(n=21, mean = 4450074, sd = 11)
    , tree_height_m = exp(rgamma(n = 21, shape = (7/4)^2, rate = (4^2)/7))
  )
```

Use the `trees_dbh()` function to estimate DBH based on tree height and tree location.

```{r, results = FALSE}
# call the function
tl_dbh <- cloud2trees::trees_dbh(tree_list = tl, crs = "32613")
```

What is this data?

```{r}
tl_dbh %>% dplyr::glimpse()
```

Let's look at the relationship between tree height and tree DBH estimated from the FIA plot data.

```{r}
tl_dbh %>%
  ggplot2::ggplot(mapping = ggplot2::aes(x = tree_height_m, y = dbh_cm)) + 
  ggplot2::geom_point(color = "navy", alpha = 0.6) +
  ggplot2::labs(x = "tree ht. (m)", y = "tree DBH (cm)") +
  ggplot2::scale_x_continuous(limits = c(0,NA)) +
  ggplot2::scale_y_continuous(limits = c(0,NA)) +
  ggplot2::theme_light()
```

We can look at this data spatially too.

```{r}
# height plot
plt_ht2 <-
  tl_dbh %>% 
  ggplot2::ggplot(mapping = ggplot2::aes(color = tree_height_m)) + 
  ggplot2::geom_sf(size = 3) + 
  ggplot2::scale_color_distiller(palette = "Oranges", name = "tree ht. (m)", direction = 1) +
  ggplot2::theme_void() +
  ggplot2::theme(
    legend.position = "top", legend.direction = "horizontal"
    , panel.border = ggplot2::element_rect(color = "black", fill = NA)
  )
# diameter plot
plt_dbh2 <-
  tl_dbh %>% 
  ggplot2::ggplot(mapping = ggplot2::aes(color = dbh_cm)) + 
  ggplot2::geom_sf(size = 3) + 
  ggplot2::scale_color_distiller(palette = "Purples", name = "tree DBH (cm)", direction = 1) +
  ggplot2::theme_void() +
  ggplot2::theme(
    legend.position = "top", legend.direction = "horizontal"
    , panel.border = ggplot2::element_rect(color = "black", fill = NA)
  )
# combine with patchwork
plt_ht2 + plt_dbh2
```

## Estimate Tree Forest Type for a Tree List{#trees_type}

If we already have a list of trees with tree coordinate, we can use the `trees_type()` function attach the tree forest type based on the spatial overlap with the [Forest Type Groups of the Continental United States](https://www.arcgis.com/home/item.html?id=10760c83b9e44923bd3c18efdaa7319d) data (Wilson 2023). This forest type group layer was developed using data from over 213,000 national forest inventory plots measured during the period 2014-2018 from the FIA program and has been aggregated to 90-meter resolution for this specific project to make the data more accessible over the entire continental US.

We just need to pass a data frame with the columns `treeID`, `tree_x`, `tree_y` to the `trees_type()` function. We can also use an `sf` class object with POINT or POLYGON geometry (see `sf::st_geometry_type()`) and the program will use the data "as-is" and only require the `treeID` column.

```{r}
# a fake tree list
tl <- dplyr::tibble(
    treeID = c(1:66)
    , tree_x = rnorm(n=66, mean = 458000, sd = 75)
    , tree_y = rnorm(n=66, mean = 4450000, sd = 75)
  )
```

Use the `trees_type()` function to extract the FIA forest type group based on tree location. If a tree overlaps with an area that is classified as "non-forest", the program will search for the nearest forest type to impute a value; we'll limit the search radius by setting the `max_search_dist_m` parameter to 88 meters.  

```{r, results = FALSE}
# call the function
tl_type <- cloud2trees::trees_type(tree_list = tl, crs = "32613", max_search_dist_m = 88)
```

The return includes our tree list with forest type data (`tree_list`) as well as the FIA Forest Types Group raster (`foresttype_rast`) of the area we searched.

```{r}
tl_type %>% names()
```

What is in the tree list data?

```{r}
tl_type$tree_list %>% dplyr::glimpse()
```

Let's look at the FIA Forest Type Group data we extracted for the tree list.

```{r}
tl_type$tree_list %>%
  sf::st_drop_geometry() %>% 
  dplyr::count(forest_type_group_code, forest_type_group)
```

We can plot our spatial tree list

```{r}
# now plot
tl_type$tree_list %>%
  ggplot2::ggplot() + 
  ggplot2::geom_sf(ggplot2::aes(color=forest_type_group), size = 3) +
  ggplot2::labs(color = "FIA forest\ntype group") +
  ggplot2::scale_color_brewer(palette = "Dark2") +
  ggplot2::theme_void() +
  ggplot2::theme(panel.border = ggplot2::element_rect(color = "black", fill = NA))
```

Let's check out the FIA Forest Types Group raster (`foresttype_rast`) of the area we searched

```{r}
r_plt <- 
  tl_type$foresttype_rast %>%
    as.data.frame(xy=T) %>% 
    dplyr::rename(f = 3) %>% 
    dplyr::mutate(f = as.factor(f)) %>% 
    ggplot2::ggplot() + 
    ggplot2::geom_tile(mapping = ggplot2::aes(x=x, y=y, fill = f)) +
    ggplot2::labs(fill = "FIA forest type\ngroup code") +
    ggplot2::scale_fill_viridis_d(option = "turbo", alpha = 0.9) +
    ggplot2::theme_void()
r_plt
```

See the [Forest Type Groups of the Continental United States](https://www.arcgis.com/home/item.html?id=10760c83b9e44923bd3c18efdaa7319d) data (Wilson 2023) for a list of possible forest type group codes

Let's overlay our tree points on the raster data

```{r}
r_plt +
  ggplot2::geom_sf(
    data = tl_type$tree_list %>%  
      # we have to reproject
      sf::st_transform(
        crs = tl_type$foresttype_rast %>%
          terra::crs(describe=T) %>%
          dplyr::pull(code) %>%
          as.numeric() %>%
          sf::st_crs()
      )
    , mapping = ggplot2::aes(shape = forest_type_group)
    , color = "white"
    , size = 2
  ) +
  ggplot2::labs(shape = "FIA forest\ntype group") +
  ggplot2::guides(shape = ggplot2::guide_legend(override.aes = list(size = 3, color = "black")))
```

## Estimate Tree CBH for a Tree List{#trees_cbh}

If you wish to estimate crown base height (CBH) as part of the point cloud processing the `LadderFuelsR` package ([https://github.com/olgaviedma/LadderFuelsR](https://github.com/olgaviedma/LadderFuelsR)) and `leafR` package ([https://github.com/DRAAlmeida/leafR](https://github.com/DRAAlmeida/leafR)) must be manually installed first.

``` r
# install.packages("remotes")
## install LadderFuelsR
remotes::install_github(repo = "olgaviedma/LadderFuelsR", upgrade = F)
## install leafR
remotes::install_github(repo = "DRAAlmeida/leafR", upgrade = F)
```

After installing these packages, if we already have spatial polygons of tree crowns and height normalized point cloud data, we can attempt to extract tree CBH from the point cloud using the `trees_cbh()` function.

We just need to pass a `sf` class object with POLYGON geometry and the columns `treeID` and `tree_height_m` and the height normalized point cloud data to the `trees_cbh()` function.

We'll use the tree crown polygons and normalized point cloud data examples that ship with the `cloud2trees` package.

```{r, results = FALSE}
# read example crown polygons
f <- system.file(package = "cloud2trees","extdata", "crowns_poly.gpkg")
p <- sf::st_read(f, quiet = T)
# path to the normalized point cloud data
nlas <- paste0(system.file(package = "cloud2trees"),"/extdata/norm_las")
# call the function
trees_cbh_ans <- cloud2trees::trees_cbh(
  trees_poly = p
  , norm_las = nlas
  , tree_sample_prop = 0.5
  , estimate_missing_cbh = TRUE
)
```

What is this data?

```{r}
trees_cbh_ans %>% 
  dplyr::select(treeID, tree_height_m, tree_cbh_m, is_training_cbh) %>% 
  dplyr::glimpse()
```

Let's look at the relationship between tree height and tree CBH as extracted from the point cloud. Note, that we do not expect a perfect linear relationship between tree height and CBH throughout the entire height range because CBH is also determined spatially (e.g. as a fire moves through a stand).

```{r}
trees_cbh_ans %>%
  dplyr::arrange(is_training_cbh) %>%
  ggplot2::ggplot(mapping = ggplot2::aes(x = tree_height_m, y = tree_cbh_m, color=is_training_cbh)) + 
  ggplot2::geom_point() +
  ggplot2::labs(x = "tree ht. (m)", y = "tree CBH (m)") +
  ggplot2::scale_y_continuous(breaks = scales::extended_breaks(n=12)) +
  ggplot2::scale_x_continuous(breaks = scales::extended_breaks(n=14)) +
  ggplot2::scale_color_viridis_d(alpha = 0.8, name = "is CBH\nfrom cloud?") +
  ggplot2::theme_light()
```

We can look at this data spatially too.

```{r}
trees_cbh_ans %>%
  dplyr::arrange(is_training_cbh) %>%
  ggplot2::ggplot(mapping = ggplot2::aes(fill = tree_cbh_m, color=is_training_cbh)) + 
  ggplot2::geom_sf() +
  ggplot2::scale_color_viridis_d(alpha = 0.8, name = "is CBH\nfrom cloud?") +
  ggplot2::scale_fill_distiller(palette = "Greens", name = "tree CBH (m)", direction = 1) +
  ggplot2::theme_void() +
  ggplot2::theme(
    legend.position = "top", legend.direction = "horizontal"
    , panel.border = ggplot2::element_rect(color = "black", fill = NA)
  ) +
  ggplot2::guides(
    color = ggplot2::guide_legend(override.aes = list(lwd = 3, fill = NA))
  )
```

## Estimate Tree HMD for a Tree List{#trees_hmd}

If you wish to extract the height of the maximum crown diameter (HMD) using height normalized point cloud data (e.g. as exported by `cloud2raster()`) the `trees_hmd()` function and the `estimate_tree_hmd` in the `cloud2trees()` function may be relevant to your interests.

We just need to pass a `sf` class object with POLYGON geometry and the columns `treeID` and `tree_height_m` and the height normalized point cloud data to the `trees_hmd()` function. The function returns the data with the added columns: `max_crown_diam_height_m`, `is_training_hmd`.

We'll use the tree crown polygons and normalized point cloud data examples that ship with the `cloud2trees` package.

```{r, results = FALSE}
# read example crown polygons
f <- system.file(package = "cloud2trees","extdata", "crowns_poly.gpkg")
p <- sf::st_read(f, quiet = T)
# path to the normalized point cloud data
nlas <- paste0(system.file(package = "cloud2trees"),"/extdata/norm_las")
# call the function
trees_hmd_ans <- cloud2trees::trees_hmd(
  trees_poly = p
  , norm_las = nlas
  , estimate_missing_hmd = TRUE
)
```

What is this data?

```{r}
trees_hmd_ans %>% 
  dplyr::select(treeID, tree_height_m, max_crown_diam_height_m, is_training_hmd) %>% 
  dplyr::glimpse()
```

Let's look at the relationship between tree height and tree HMD as extracted from the point cloud. Note, that we do not expect a perfect linear relationship between tree height and HMD throughout the entire height range because HMD is also determined spatially (e.g. as a fire moves through a stand).

```{r}
trees_hmd_ans %>%
  dplyr::arrange(is_training_cbh) %>%
  ggplot2::ggplot(
    mapping = ggplot2::aes(x = tree_height_m, y = max_crown_diam_height_m, color=is_training_hmd)
  ) + 
  ggplot2::geom_point() +
  ggplot2::labs(x = "tree ht. (m)", y = "tree HMD (m)") +
  ggplot2::scale_y_continuous(breaks = scales::extended_breaks(n=12)) +
  ggplot2::scale_x_continuous(breaks = scales::extended_breaks(n=14)) +
  ggplot2::scale_color_viridis_d(option = "turbo", begin = 0.2, alpha = 0.8, name = "is HMD\nfrom cloud?") +
  ggplot2::theme_light()
```

We can look at this data spatially too.

```{r}
trees_hmd_ans %>%
  dplyr::arrange(is_training_hmd) %>%
  ggplot2::ggplot(mapping = ggplot2::aes(fill = max_crown_diam_height_m, color=is_training_hmd)) + 
  ggplot2::geom_sf() +
  ggplot2::scale_color_viridis_d(option = "turbo", begin = 0.2, alpha = 0.8, name = "is HMD\nfrom cloud?") +
  ggplot2::scale_fill_distiller(palette = "Greys", name = "tree HMD (m)", direction = 1) +
  ggplot2::theme_void() +
  ggplot2::theme(
    legend.position = "top", legend.direction = "horizontal"
    , panel.border = ggplot2::element_rect(color = "black", fill = NA)
  ) +
  ggplot2::guides(
    color = ggplot2::guide_legend(override.aes = list(lwd = 3, fill = NA))
  )
```

## Estimate Tree Biomass for a Tree List{#trees_biomass}

The `cloud2trees()` package includes methods for estimating individual tree biomass in kilograms, or the component biomass of the tree crown in kilograms.

Currently supported methods for estimating biomass include:

* Tree crown biomass in kilograms:
  - "landfire" - based on [LANDFIRE's Forest Canopy Bulk Density (CBD)](https://landfire.gov/fuel/cbd) data (`trees_biomass_landfire()`)
  - "cruz" - based on [Cruz et al. (2003)](https://scholar.google.com/scholar?cluster=316241498622221569&oi=gsb&hl=en&as_sdt=0,5) canopy fuel stratum equations (`trees_biomass_cruz()`)

The `trees_biomass()` function streamlines the process for estimating individual tree biomass in kilograms, or the component biomass of the tree crown in kilograms. Users can select one, some, or all of the methods available in the package for estimating biomass.

We just need to pass a data frame with the columns `treeID`, `tree_x`, `tree_y` to the `trees_biomass*()` function. We can also use an `sf` class object with POINT or POLYGON geometry (see `sf::st_geometry_type()`) and the program will use the data "as-is" and only require the `treeID` column.

Since estimating tree crown biomass using stand-based fuel estimates requires back-transformation to the tree level, the `trees_biomass_landfire()`, `trees_biomass_cruz()`, and `trees_biomass()` if using the Cruz or LANDFIRE methods also require the following data attributes:

* `crown_area_m2`, `tree_height_m` (e.g. as exported by `raster2trees()`)
* `tree_cbh_m` (e.g. as exported by `trees_cbh()`)
* and one of `dbh_cm`, `dbh_m`, or  `basal_area_m2` (e.g. as exported by `trees_dbh()`)

We'll use the tree crown polygons that ship with the `cloud2trees` package.

```{r, results = FALSE}
# read example crown polygons
f <- system.file(package = "cloud2trees","extdata", "crowns_poly.gpkg")
tl <- sf::st_read(f, quiet = T)
```

### `trees_biomass()`

The `trees_biomass()` function streamlines the process for estimating individual tree biomass in kilograms, or the component biomass of the tree crown in kilograms. Users can select one, some, or all of the methods available in the package for estimating biomass. The following function calls are equivalent:

``` r
# trees_biomass with method = "landfire"
trees_biomass(tree_list, method = "landfire")
# is equivalent to
trees_biomass_landfire(tree_list)
```

We'll use the `trees_biomass()` function and ask for both the LANDFIRE and the Cruz et al. (2003) estimates of tree crown biomass in kilograms with the argument `method = c("landfire","cruz")`

```{r, warning=FALSE, results=F}
# call trees_biomass and get multiple biomass estimates
trees_biomass_ans <- trees_biomass(tree_list = tl, method = c("landfire","cruz"))
```

what did we get back?

```{r}
trees_biomass_ans %>% names()
```

check out the tree list data

```{r}
trees_biomass_ans$tree_list %>% dplyr::glimpse()
```

that's a lot of extra information...the `cruz_crown_biomass_kg` and `landfire_crown_biomass_kg` columns include estimates of tree crown biomass in kilograms that we are after

plot tree LANDFIRE and Cruz crown biomass estimate

```{r}
library(patchwork)
# plot tree landfire crown biomass estimate
p1 <- trees_biomass_ans$tree_list %>%
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = tree_height_m
      , y = landfire_crown_biomass_kg
      , color = crown_area_m2
    )
  ) +
  ggplot2::geom_point()
# plot tree cruz crown biomass estimate
p2 <- trees_biomass_ans$tree_list %>%
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = tree_height_m
      , y = cruz_crown_biomass_kg
      , color = crown_area_m2
    )
  ) +
  ggplot2::geom_point()
# patchwork it
p1/p2
```

the estimates look similar but not exactly the same. let's plot them against each other

```{r}
# get the max for the upper limit scale
ul <- max(
  trees_biomass_ans$tree_list$cruz_crown_biomass_kg
  , trees_biomass_ans$tree_list$landfire_crown_biomass_kg
)
# plot tree landfire vs. cruz crown biomass estimate
trees_biomass_ans$tree_list %>%
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = landfire_crown_biomass_kg, y = cruz_crown_biomass_kg
    )
  ) +
  ggplot2::geom_abline(lwd = 1.5) +
  ggplot2::geom_smooth(method = "lm", se=F, color = "gray", linetype = "dashed") +
  ggplot2::geom_point(ggplot2::aes(color = tree_height_m)) +
  ggplot2::scale_x_continuous(limits = c(0, ul)) +
  ggplot2::scale_y_continuous(limits = c(0, ul))
```

let's check out the LANDFIRE stand data

```{r}
trees_biomass_ans$stand_cell_data_landfire %>% dplyr::filter(trees>0) %>% dplyr::glimpse()
```

this data includes the cell data (where, a "stand" is represented by a raster cell) from the LANDFIRE Forest Canopy Bulk Density (CBD) data

we can use this stand/cell data as raster data and overlay the tree points...let's do this for the LANDFIRE data

```{r}
# get the projection for the stand cell data
epsg_code <- trees_biomass_ans$stand_cell_data_landfire$rast_epsg_code[1] %>% as.numeric()
# plot the stand cell data with trees overlaid
trees_biomass_ans$stand_cell_data_landfire %>%
  dplyr::filter(trees>0) %>%
  ggplot2::ggplot() +
  ggplot2::geom_tile(ggplot2::aes(x=x,y=y,fill = landfire_stand_kg_per_m3), color = "gray44") +
  ggplot2::geom_text(ggplot2::aes(x=x,y=y,label = trees), color = "white") +
  ggplot2::geom_sf(
    data = trees_biomass_ans$tree_list %>% sf::st_transform(crs = epsg_code)
    , ggplot2::aes(color = cruz_crown_biomass_kg)
  ) +
  ggplot2::labs(fill="landfire\nstand kg/m3", color = "landfire\ncrown kg", caption = "# trees shown in cell") +
  ggplot2::scale_fill_viridis_c(option = "rocket", na.value = "gray", direction = -1) +
  ggplot2::scale_color_viridis_c(option = "viridis", na.value = "gray22", begin = 0.6) +
  ggplot2::theme_void()
```

