---
title: "Overview"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE
, comment = "#>"
, out.width = "100%"
, echo = TRUE
, warning = FALSE
, message = FALSE
)
library(rgl)
rgl::setupKnitr(autoprint = TRUE)
options(rgl.useNULL = TRUE) # Suppress the separate window when knitting
```

The cloud2trees package offers compiled processing routines for generating individual-tree forest inventories directly from aerial point cloud data. Designed to integrate easily into existing silviculture and fire management workflows, the package streamlines complex processing steps that are often fragmented across multiple software tools. 

This demonstration provides a broad overview of cloud2trees by walking through the entire process, starting with an overview of point cloud data, progressing the data through the core tree detection and metric extraction pipeline, and concluding with visualization and aggregation of the final outputs. The goal is to illustrate how the package quickly transforms raw 3D data into actionable, attributable, and spatially explicit tree lists ready for further analysis or planning and modeling applications.

Let's load the libraries we'll use

```{r}
library(cloud2trees)
library(ggplot2)
library(magrittr)
library(terra)
library(sf)
library(lidR)
```

## Raw Point Cloud Data

We'll start with a quick overview of aerial point cloud data and its typical processing in R, independent of the cloud2trees package. Aerial point cloud data are highly detailed 3D datasets that capture the forest's structure. This data is typically acquired via fixed wing crewed aircraft or uncrewed aerial systems (UAS) equipped with lidar sensors or high-resolution cameras for Digital Aerial Photogrammetry (DAP). We will demonstrate basic handling using the widely adopted `lidR` package, including initial file loading (`.las`/`.laz`), metadata review, and essential 3D visualization. See the `lidR` [book](https://r-lidar.github.io/lidRbook) by point cloud processing expert [Jean-Romain Roussel](https://github.com/Jean-Romain) for excellent detail on data processing with this foundational package.

While `lidR` provides the necessary foundational tools for these tasks, cloud2trees steps in to fill the crucial gap by compiling and automating the more complex, sequential processing steps, including built-in handling for large extent data that can be challenging for users scripting a pipeline from scratch.

Let's load the data for the tutorial which is a small data set that ships with the `lidR` package

```{r}
# the path to a single .las file
las_fpath <- system.file(package = "lidR", "extdata", "MixedConifer.laz")
# load the single file point cloud with lidR selecting only the primary information
las_data <- lidR::readLAS(las_fpath, select = "xyzic")
```

what is this data?

```{r}
las_data
```

that is very useful information and we can explore the actual data of the point cloud to get more detail

```{r}
las_data@data %>% dplyr::glimpse()
```

we can explore the X, Y, and Z data further

```{r}
las_data@data %>% dplyr::select(X,Y,Z) %>% summary()
```

because this is a relatively small data set, we can visualize the 3D data with the points colored by the Z measurement

```{r}
lidR::plot(
  x = las_data
  , color = "Z", bg = "white"
  , legend = T
  , pal = grDevices::heat.colors(55)
  , rgl = T
)
rgl::rglwidget()
```

now that we've covered the foundational concepts of point cloud data, we're ready to put our cloud2trees pipeline into action using the core `cloud2trees()` function to rapidly generate a complete forest inventory.

## Core `cloud2trees()` Pipeline

The heart of the package is the `cloud2trees()` function, which runs the entire Individual Tree Detection (ITD) and attribute extraction workflow based on settings defined by the user. To illustrate its immediate simplicity, we will first demonstrate the execution of this all-in-one pipeline using the default settings, specifying just the input location of the example point cloud and where we want to save the results (a temporary directory for this example). While we strongly recommend that users customize the function parameters to meet their specific study objectives, this demonstration will illustrate how simply cloud2trees can process raw point cloud data to produce a spatial forest inventory tree list as well as intermediate products like the Digital Terrain Model (DTM) and Canopy Height Model (CHM).

```{r, results=F, warning=F, message=F}
cloud2trees_ans <- cloud2trees::cloud2trees(
  input_las_dir = las_fpath
  , output_dir = tempdir()
)
```

That single `cloud2trees()` function call is all that is needed to produce a spatial forest inventory tree list, DTM, and CHM from raw point cloud data

## Reviewing Outputs

Regardless of the custom parameters selected by the user, upon successful completion over an area with trees, the `cloud2trees()` function will always return a spatial forest inventory tree list (including core attributes such as height, location, and crown area as is typical of ITD processing), as well as the intermediate DTM and CHM rasters. This section focuses on reviewing these results, visualizing the segmented crowns and tree tops.

Let's check out what is included in the return from the `cloud2trees()` function.

```{r}
# what is it?
cloud2trees_ans %>% names()
```

There is a digital terrain model (DTM) raster 

```{r}
# there's a DTM
cloud2trees_ans$dtm_rast
```

The DTM was created at the `cloud2trees()` default `r max(terra::res(cloud2trees_ans$dtm_rast))` m resolution and that the elevations vary between `r scales::comma(terra::minmax(cloud2trees_ans$dtm_rast)[1],accuracy = 0.01)` and `r scales::comma(terra::minmax(cloud2trees_ans$dtm_rast)[2],accuracy = 0.01)` m.

we can plot the DTM using `terra::plot()`

```{r}
terra::plot(cloud2trees_ans$dtm_rast)
```

While the plot shows a fair amount of variation, let’s remember it only represents `r scales::comma(diff(terra::minmax(cloud2trees_ans$dtm_rast)),accuracy = 0.01)` m of vertical relief. If we looked at an area with 60 m of vertical relief, for example, we would not even notice these small fluctuations.

There is a canopy height model (CHM) raster 

```{r}
# there's a CHM
cloud2trees_ans$chm_rast
```

The CHM was created at the `cloud2trees()` default `r max(terra::res(cloud2trees_ans$chm_rast))` m resolution and that the elevations vary between `r scales::comma(terra::minmax(cloud2trees_ans$chm_rast)[1],accuracy = 0.01)` and `r scales::comma(terra::minmax(cloud2trees_ans$chm_rast)[2],accuracy = 0.01)` m.

we can plot the CHM using `terra::plot()`

```{r}
terra::plot(cloud2trees_ans$chm_rast, col = grDevices::heat.colors(55, alpha = 0.88))
```

the `treetops_sf` object in the return from `cloud2trees()` is a spatial data frame representing the extracted tree tops as individual points

```{r}
cloud2trees_ans$treetops_sf %>% dplyr::glimpse()
```

The tree list contains `r nrow(cloud2trees_ans$treetops_sf)` rows and `r ncol(cloud2trees_ans$treetops_sf)` columns. The `r nrow(cloud2trees_ans$treetops_sf)` rows means that `cloud2trees()` detected `r nrow(cloud2trees_ans$treetops_sf)` trees across this `r terra::expanse(cloud2trees_ans$dtm_rast,unit = "m")[2] %>% as.numeric() %>% scales::comma(scale = 1e-4,accuracy = 0.01,suffix = " hectare")` area. The default use of `cloud2trees()` provides the height, crown area, and location (X and Y coordinate) for each of the trees it identified. However, we also see there are many columns with `NA` values and we will explore how cloud2trees can estimate these attributes in later tutorials.

the `crowns_sf` object in the return from `cloud2trees()` is a spatial data frame representing the extracted tree crowns as polygons

```{r}
cloud2trees_ans$crowns_sf %>% dplyr::glimpse()
```

Notice that `cloud2trees_ans$crowns_sf` and `cloud2trees_ans$treetops_sf` have the *exact same structure* but one is spatial polygons and the other is spatial points.

Now let's create a visual of the individual tree crowns stored in the `crowns_sf` object overlaid on the CHM. This time, we'll plot using the `ggplot2` package

```{r}
cloud2trees_ans$chm_rast %>%
  terra::as.data.frame(xy = T) %>%
  dplyr::rename(f = 3) %>%
  ggplot2::ggplot() +
  ggplot2::geom_tile(
    mapping = ggplot2::aes(x = x, y = y, fill = f)
    ) +
  ggplot2::geom_sf(
    data = cloud2trees_ans$crowns_sf
    , color = "grey33"
    , lwd = 0.8
    ) +
  ggplot2::scale_fill_gradientn(colors = grDevices::heat.colors(55, alpha = 0.88)) +
  ggplot2::labs(fill = "CHM (m)") +
  ggplot2::theme_void()
```

This figure displays the individual tree crowns that `cloud2trees()` identified using its default variable window function for tree detection (i.e. the `ws` argument). If you look closely, you can likely find places where tree crowns are being over or under divided. In a later tutorial we will look at how the `itd_tuning()` function of cloud2trees can be used to optimize local tree detection given the crown architecture of the forest being analyzed.

## Analyzing Outputs

The ultimate goal of cloud2trees is to provide accurate and spatially explicit inputs that directly support management decisions. This final section demonstrates how the output spatial tree list can be immediately used for forest management and planning. We will showcase examples such as calculating stand-level metrics (e.g. trees per hectare and mean stand height), generating height distributions, and identifying potentially ecologically important trees. This brief overview is meant to showcase how easily insight can be gained from the `cloud2trees()` spatial inventory, improving access to data and enabling cross-disciplinary integration to spur novel forest management approaches and enable collaborative action.

### Stand-Level Metrics

We'll pretend our forest stand is the central 4,000 m^2^ of the point cloud data and create a square polygon to use as our stand boundary

```{r}
my_stand <-
  cloud2trees_ans$treetops_sf %>% 
    sf::st_union() %>% 
    sf::st_centroid() %>% 
    sf::st_buffer(sqrt(4000)/2, endCapStyle = "SQUARE")
```

Let's look the tree top points identified by `cloud2trees()` in relation to the stand boundary. We'll color the tree top points by tree height

```{r}
# what is this?
ggplot2::ggplot() + 
  # tree tops
  ggplot2::geom_sf(
    data = cloud2trees_ans$treetops_sf
    , mapping = ggplot2::aes(color = tree_height_m)
    , size = 1.8
  ) + 
  # stand polygon
  ggplot2::geom_sf(
    data = my_stand
    , fill = NA, color = "black", lwd = 1.1
  ) +
  ggplot2::scale_color_distiller(
    palette = "Blues", direction = 1, name = "tree ht. (m)"
  ) +
  ggplot2::theme_void()
```

to get stand-level metrics we can crop our tree list to the stand boundary and summarize. we'll do this all in one tidy pipeline but you can break the pipes (`%>%`) to see what each step does. 

```{r}
# first we'll crop the tree list to the stand
treetops_in_stand <- cloud2trees_ans$treetops_sf %>% sf::st_intersection(my_stand)
# summarize
stand_summary <- 
  treetops_in_stand %>% 
  sf::st_drop_geometry() %>% # don't need geom now
  # summarize
  dplyr::summarise(
    mean_tree_height_m = mean(tree_height_m)
    , n_trees = dplyr::n()
  ) %>% 
  # add stand information
  dplyr::mutate(
    stand_area_m2 = sf::st_area(my_stand) %>% as.numeric()
    , stand_area_ha = stand_area_m2/10000
    , trees_per_ha = n_trees/stand_area_ha
  )
# what is this?
stand_summary
```

`cloud2trees()` identified `r scales::comma(stand_summary$n_trees,accuracy=1)` trees in the `r scales::comma(stand_summary$stand_area_ha,accuracy=0.01)` ha stand resulting in `r scales::comma(stand_summary$trees_per_ha,accuracy=0.1)` trees per hectare (TPH). Based on these `r scales::comma(stand_summary$n_trees,accuracy=1)` trees, the mean stand height is `r scales::comma(stand_summary$mean_tree_height_m,accuracy=0.1)` m.

### Height Distributions

We'll continue to look at the trees within our stand to explore the tree height distribution.

A simple density plot of heights can be generated quickly

```{r}
treetops_in_stand %>% 
  ggplot2::ggplot() + 
  ggplot2::geom_density(
    ggplot2::aes(x = tree_height_m)
    , alpha = 0.88
    , fill = "navy", color = NA
  ) +
  ggplot2::labs(x = "tree ht. (m)") +
  ggplot2::theme_light()
```

we can also create custom height bins and summarize the trees within those bins. again, this is one big tidy pipeline but you can break the pipes (`%>%`) to see what each step does.

```{r}
treetops_in_stand %>%
  sf::st_drop_geometry() %>%
  dplyr::mutate(
    height_bin = ggplot2::cut_width(
      tree_height_m, width = 3, boundary = 0, closed = "left"
    )
  ) %>% 
  dplyr::group_by(height_bin) %>% 
  dplyr::summarise(
    n_trees = dplyr::n()
  ) %>%
  dplyr::ungroup() %>% 
  dplyr::mutate(
    stand_area_m2 = sf::st_area(my_stand) %>% as.numeric()
    , stand_area_ha = stand_area_m2/10000
    , trees_per_ha = n_trees/stand_area_ha
    , tot_trees_per_ha = sum(trees_per_ha)
    , pct = trees_per_ha/tot_trees_per_ha
    , height_bin_lab = paste0(
      stringr::word(
          height_bin
          , 1
          , sep = stringr::fixed(",")
        ) %>% readr::parse_number()
      , " to "
      , stringr::word(
          height_bin
          , -1
          , sep = stringr::fixed(",")
        ) %>% readr::parse_number() %>% `-`(0.1)
    ) %>% 
    factor() %>% 
    forcats::fct_reorder(
      stringr::word(
        height_bin
        , 1
        , sep = stringr::fixed(",")
      ) %>% readr::parse_number()
    )
  ) %>% 
  # plot
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = height_bin_lab, y = trees_per_ha
      , fill = trees_per_ha
      , label = paste0(
        scales::comma(trees_per_ha, accuracy = 0.1)
        , "\n"
        , scales::percent(pct, accuracy = 0.1)
      )
    )
  ) +
  ggplot2::geom_col(width = 0.7) +
  ggplot2::geom_text(color = "black", size = 3, vjust = -0.2) +
  ggplot2::scale_fill_distiller(palette = "Purples", direction = 1) +
  ggplot2::scale_y_continuous(
    labels = scales::comma_format(accuracy = 1)
    , expand = ggplot2::expansion(mult = c(0, 0.1))
  ) +
  ggplot2::labs(
    x = "height (m) class"
    , y = "TPH"
    , title = "TPH by height class"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , axis.text.y = ggplot2::element_text(size = 10)
    , axis.text.x = element_text(angle = 90, size = 10, vjust = 0.5, hjust = 1)
  )
```

### Identify Potentially Ecologically important trees

finally, let's identify potentially ecologically important trees based on a height threshold of 24 m across the entire extent of the point cloud data

```{r}
cloud2trees_ans$treetops_sf %>% 
  dplyr::mutate(
    tall_tree = ifelse(tree_height_m>=24,"Tall Tree", "Other")
  ) %>% 
  ggplot2::ggplot() + 
  # tree tops
  ggplot2::geom_sf(
    mapping = ggplot2::aes(color = tall_tree)
    , size = 1.8
  ) + 
  # stand polygon
  ggplot2::geom_sf(
    data = my_stand
    , fill = NA, color = "black", lwd = 1.1
  ) +
  ggplot2::scale_color_viridis_d(option = "viridis", direction = -1, name = "") +
  ggplot2::theme_void() +
  ggplot2::theme(legend.position = "top") +
  ggplot2::guides(color = ggplot2::guide_legend(override.aes = list(size = 5)))
```

since we have the X and Y coordinates of all trees, we can easily load them onto a tablet or GPS device to navigate to those important trees in the field
